|

### 2025-05-29 04-discussion.Rmd
edits Discussion Vicky

The current results indicate several important implications. First, we found that students significantly benefit from AI-generated step-by-step reasoning accompanying solutions, mainly when performing open-ended tasks like estimating unknown quantities. This finding aligns with previous research demonstrating that LLMs can enhance student performance and that explanations help children develop critical thinking. However, our results underscore that the key mechanism driving AI improvements is the step-by-step reasoning provided by the AI, which allows students to understand better and engage with problem-solving. This insight suggests that teachers should focus on educating students to enhance their ability to evaluate and judge the correctness of AI-generated reasoning.

Furthermore, our study reveals that step-by-step reasoning not only helps students in solving problems but also enhances students' ability to evaluate AI-generated information critically. This was evidenced by the more positive evaluations of human-generated guesses compared to AI-generated solutions, suggesting that students could better assess and challenge AI predictions. This contribution extends the existing literature by showing that AI can foster critical thinking and analytical skills when coupled with transparent reasoning, instead of being presented  as a "black box."

In addition, we compared various ways to structure student-AI interactions and, more specifically, the effectiveness of Socratic AI (an interactive, questioning-based AI) with non-Socratic AI. Our results showed that Socratic AI was more engaging and promoted greater interaction, aligning with the notion that AI-student interactions should be dynamic and dialogical. However, contrary to our expectations, we found no significant differences in students' self-reported confidence in the accuracy of their answers or in the correctness of their responses despite the more frequent interactions with the Socratic AI. Additionally, the Socratic AI's perceived helpfulness was rated lower compared to the non-Socratic AI. These results cast some doubts on the effectiveness of Socratic AI in short-term tasks or, more broadly, the effect of certain kinds of AI-student interactions. As such, existing pedagogical practices extensively used in human-human interaction might not always work in human-AI interaction. This underscores that new pedagogical paradigms are needed to integrate AI into pedagogical practices effectively.

Our findings indicate that simply interacting with AI cannot promote meaningful and lasting learning. In our initial test on sound propagation, approximately half of the students could revise their initial answers based on AI interactions, improving their response accuracy from 30% to 70%. However, in a verification task where students had no access to AI, the majority failed to identify the correct answer, mistakenly claiming that sound propagates faster in water than in gold. Most students exhibited this misunderstanding, which supports key concerns that AI-generated answers alone may not facilitate effective learning, and that Socratic AI does not mitigate this risk.

Our results suggest that AI has great potential as an educational tool, but its implementation requires careful consideration. First, students with prior experience in using AI tools may not readily adopt new pedagogical approaches, especially if they perceive them as less effective than commercially available alternatives. Second, the effectiveness of the pedagogical approach may vary depending on the nature of the task, complicating the design of a one-size-fits-all solution for AI-assisted learning.

@@

@@

While our study focused on comparing a fairly general pedagogical approach---Socratic AI, future research should explore the effectiveness of alternative pedagogical approaches and the scalability of AI tools across diverse educational contexts. Yet, our findings have important implications for designing AI tutors, highlighting the importance of  providing transparent AI-generated step-by-step reasoning and the challenges of fostering learning through guided AI-student interactions. Therefore, our study suggests that AI systems must engage students interactively and foster critical thinking and problem-solving skills to maximise their educational value. Future developments should focus on refining the integration of AI-generated reasoning and ensuring that AI tools are adaptable to various learning tasks and students' needs.

>>>>

The present study aims to contribute to our understanding of the impact of pedagogically aligned configurations of an LLM-based tool on high-school students’ learning and attitudes with a special focus on their critical thinking.

With the rapid adoption of AI in education, the focus on students' competencies and skills, such as critical thinking, has become particularly urgent [@walter2024embracing;@holmes2023guidance]. Despite the rapidly increasing body of research on the impact of LLMs on students’ learning, most studies focus on participants in higher education [e.g.,@suriano2025student], which do not necessarily generate insights that apply to younger students who have different cognitive maturity. In addition, despite the increasing research interest regarding the impact of LLMs on students’ learning, there is little consensus among researchers, especially due to students’ over-reliance on AI tools [@zhai2024effects].

<!-- as individuals increasingly favor fast and optimal solutions over slow ones constrained by practicality -->

The results of this study focus on the role of two interventions on the configuration of an LLM chatbot for educational support: (i) step-by-step explanations and (ii) the Socratic method with guiding questions. First, we found that students significantly benefit from AI-generated step-by-step reasoning accompanying solutions, mainly when performing open-ended tasks like estimating unknown quantities. This finding aligns with expectations of current research on the use of LLM chatbots and their ability to provide explanations for learning purposes [@wu2024critical]. It also aligns with previous research demonstrating that LLMs can enhance student performance [@xing2025development]. However, our results underscore that the key mechanism driving AI improvements is the step-by-step reasoning provided by the AI, which allows students to understand better and engage with problem-solving.

<!-- This insight suggests that teachers should focus on educating students to enhance their ability to evaluate and judge the correctness of AI-generated reasoning. -->

Furthermore, our study reveals that step-by-step reasoning not only helps students in solving problems but also enhances students' ability to evaluate AI-generated information critically. Critical thinking is indeed a complex cognitive process that involves evaluation and analysis of a given information [@lai2011critical]. The mobilisation of students' critical thinking in the specific task was evidenced by the more positive evaluations of human-generated guesses compared to AI-generated solutions, suggesting that students could better assess and challenge AI predictions. This contribution extends the existing literature by showing that AI can foster critical thinking and analytical skills when coupled with transparent reasoning, instead of being presented  as a "black box" [@bearman2023learning].

In addition, we compared various ways to structure student-AI interactions and, more specifically, the effectiveness of Socratic AI (an interactive, questioning-based AI) with non-Socratic AI. Our results showed that Socratic AI was more engaging and promoted greater interaction, aligning with the notion that AI-student interactions should be dynamic and dialogical [@suriano2025student]. However, contrary to our expectations, we found no significant differences in students' self-reported confidence in the accuracy of their answers or in the correctness of their responses despite the more frequent interactions with the Socratic AI. Additionally, the Socratic AI's perceived helpfulness was rated lower compared to the non-Socratic AI. These results cast some doubts on the effectiveness of Socratic AI in short-term tasks or, more broadly, the effect of certain kinds of AI-student interactions. As such, existing pedagogical practices extensively used in human-human interaction might not always work in human-AI interaction. In addition, while recent research attempts to understand the quality of Socratic LLMs for critical thinking [e.g.,@favero2024enhancing], these studies use simulations without experiments with human participants. Our results underscore the need for adapting existing pedagogical paradigms or proposing new ones for integrating LLM tools into pedagogical practices effectively.

Our findings also indicate that simply interacting with AI cannot promote meaningful and lasting learning. In our initial test on sound propagation, approximately half of the students could revise their initial answers based on AI interactions, improving their response accuracy from 30% to 70%. However, in a verification task where students had no access to AI, the majority failed to identify the correct answer, mistakenly claiming that sound propagates faster in water than in gold. Most students exhibited this misunderstanding, which supports key concerns that AI-generated answers alone do not scaffold effective learning, and the mere implementation of Socratic AI does not mitigate this risk.

Our results suggest that AI has great potential as an educational tool, but its implementation requires careful considerations. First, students with prior experience in using AI tools may not readily adopt new pedagogical approaches, especially if they perceive them as less effective than commercially available alternatives. Second, the effectiveness of the pedagogical approach may vary depending on the nature of the task, complicating the design of a one-size-fits-all solution for AI-assisted learning.

@@

Our future work includes a large-scale study with a representative sample in a country in Europe, where we aim to address the above-mentioned limitations.

@@

While our study focused on comparing a fairly general pedagogical approach---Socratic AI, future research should explore this pedagogical method in a more nuanced way and consider alternative pedagogical approaches to facilitate scalability of AI tools across diverse tasks and educational contexts. Yet, our findings have important implications for designing AI tutors, highlighting the importance of  providing transparent AI-generated step-by-step reasoning and the challenges of fostering learning through guided AI-student interactions. Therefore, AI systems must engage students interactively and provide learning opportunities for students' critical thinking and problem-solving skills to maximise their educational value. Future developments should focus on refining the integration of AI-generated reasoning and ensuring that AI tools are adaptable to various learning tasks and students' needs.
